---
title: "Estimating Acreage of Nonnative Species By Borough"
output: html_document
date: "2024-03-21"
---

```{r setup, include=FALSE}
library(caret)
library(sf)
library(magrittr)
library(dplyr)
library(terra)
library(stars)
library(tmap)
library(htmlwidgets)
library(here)
library(basemaps)
library(tidyr)

here::i_am('R/classify_nonnative_species_nyc.Rmd')
knitr::opts_chunk$set(echo = TRUE)


```

### Create raster stack of input variables cropped to park boundaries

```{r}
# park boundaries
pk <- read_sf(here('data','parks.shp')) %>%
  st_transform(26918)

# boro boundaries
b <- read_sf(here('data','bb.shp')) %>%
  filter(boro_name=='Staten Island') %>% ### specify boro you want
  st_transform(26918)

# filter for parks in given borough
pk1 <- st_filter(pk,b,.predicate=st_intersects)

metrics <- read_stars('si_metrics.tif') %>%
  split()

bands <- read_stars(file.path(here() %>% dirname(),'thesis','data','nyc_data','scheme_2','si_scheme2.tif')) %>%
  split()

names(metrics) <- c('sos','eos','los1','los2','los3','peak','pop','av_greenup_diff','av_sen_diff','spring_amp','fall_amp','spring_sd','fall_sd','spring_q1','spring_q2','spring_q3','fall_q1','fall_q2','fall_q3')


bands <- bands[c(1:10,21:30),,]
names(bands) <- c('s.blue','s.green','s.red','s.re1','s.re2','s.re3','s.nir','s.nir08','s.swir1','s.swir2','f.blue','f.green','f.red','f.re1','f.re2','f.re3','f.nir','f.nir08','f.swir1','f.swir2')

bands.w <- st_warp(bands,metrics)

stack <- c(bands.w,metrics)

stack <- st_crop(stack,pk1)

```

### Create training dataset

```{r}
df <- read.csv(here('R','metrics_springfallbands.csv'))

df.na.omit <- df[rowSums(is.na(df))<39,]

nonnative <- filter(df.na.omit, inv==3|inv==4) %>%
  mutate(type='nonnative') %>%
  select(-c('inv','id','species'))

native <- filter(df.na.omit, inv==1) %>%
  mutate(type='native') %>%
  select(-c('inv','id','species'))

df2 <- rbind(nonnative,native) %>%
  mutate(type= as.factor(type))
```

### Function to train random forest model according to resampling method and class probability

```{r}

fit_model <- function(m){

  if(m=='prob'){
  fit <- train(type~.,data=Train,method='ranger',
               trControl=trainControl(method='repeatedcv',number=10,repeats=10,sampling='up',classProbs=TRUE,summaryFunction = twoClassSummary))
  }else if(m=='up'){
  ## up-sample         
  fit <- train(type~.,data=Train,method='ranger',
               trControl=trainControl(method='repeatedcv',number=10,repeats=10,sampling='up'))
  }else if(m=='down'){
  ## down-sample
  fit <- train(type~.,data=Train,method='ranger',
               trControl=trainControl(method='repeatedcv',number=10,repeats=10,sampling='down'))
  }
  return(fit)
}              
```

### Train model five times and write estimated acres of nonnative species to dataframe

```{r}
# create empty dataframe to hold results
count.df <- as.data.frame(matrix(ncol=5,nrow=0))
names(count.df) <- c('boro','native.30','nonnative.30','native.60','nonnative.60')

boro <- 'Bronx'
i <- 1
for(i in 1:5){
train.index <- createDataPartition(df2$type,p=.7,list=FALSE,times=1)

  Train <- df2[ train.index,]
  Test  <- df2[-train.index,]
  
  preprocess = preProcess(Train, method = "bagImpute")
  Train <- predict(preprocess, Train)
  Test <- predict(preprocess, Test)
  
fit <- fit_model(m='up')  # use upsampling

pred <- terra::predict(stack, fit$finalModel,
                              type = "response",
                              fun = function(model, ...) predict(model, ...)$nonnative)

# set name of attribute
names(pred) <- c('Species Type')

if(i==1){
ca <- read_stars(file.path(here() %>% dirname(),'thesis','data','nyc_data','LandsatGrid_PropCanopy2017.tif')) %>%
  st_transform(26918) %>%
  st_crop(b)

ca <- st_warp(ca,pred)
}
pred[ca<0.3] <- NA  # mask pixels < 30% tree canopy
native.30 <- 900*(table(pred$`Species Type`)[1])/4047
nonnative.30 <- 900*(table(pred$`Species Type`)[2])/4047

pred[ca<0.6] <- NA  # mask pixels < 60% tree canopy
native.60 <- (900*(table(pred$`Species Type`)[1]))/4047
nonnative.60 <- (900*(table(pred$`Species Type`)[2]))/4047

new_row <- c(boro,native.30,nonnative.30,native.60,nonnative.60)
counts.df <- rbind(counts.df,new_row)
i <- i+1
}

# summarize mean and standard devation of model results

counts.df <- as.data.frame(apply(counts.df[,2:5],2,as.numeric))
counts.df$boro <- c(rep('Queens',5),rep('Manhattan',5),rep('Brooklyn',5),rep('Bronx',5),rep('Staten Island',5))


counts.summary <- counts.df %>%
  group_by(boro) %>%
  summarize(mean.native.30 = mean(native.30),
         mean.nonnative.30 = mean(nonnative.30),
         mean.native.60 = mean(native.60),
         mean.nonnative.60 = mean(nonnative.60),
         sd.native.30 = sd(native.30),
         sd.nonnative.30 = sd(nonnative.30),
         sd.native.60 = sd(native.60),
         sd.nonnative.60 = sd(nonnative.60))

#write.csv(counts.summary,here('R','output','summary_acres_binary.csv'),row.names=FALSE)
```
### Plot results of binary model to static map


```{r}
fit <- fit_model(m='up')
pred <- terra::predict(stack, fit$finalModel,
                              type = "response",
                              fun = function(model, ...) predict(model, ...)$nonnative)

# set name of attribute
names(pred) <- c('Species Type')


ca <- read_stars(file.path(here() %>% dirname(),'thesis','data','nyc_data','LandsatGrid_PropCanopy2017.tif')) %>%
  st_transform(26918) %>%
  st_crop(b)

ca <- st_warp(ca,pred)


bm <- basemap_stars(ext=b,map_service='carto',map_type='light')

t <- tm_shape(bm) +
  tm_rgb() +
  tm_shape(pred) + 
  tm_raster(palette=c('#9BCF53','#FF204E')) +
  tm_layout(legend.outside=FALSE,
            legend.frame=TRUE,
            legend.position = c('left','top'),
            main.title=boro,
            main.title.position='center',
            main.title.size=1.2)

t

#tmap_save(t,here('R','output','bx_map.png'))

```

### Plot results of probability model to leaflet map

```{r}
fit <- fit_model(m='prob') # classProbs = TRUE

pred <- terra::predict(stack, fit$finalModel,
                              type = "response",
                              fun = function(model, ...) predict(model, ...)$nonnative)

names(pred) <- c('Native','Nonnative')


pred[ca<0.6] <- NA # mask pixels < 60% tree canopy

pl2 <- c('#4dac26','#b8e186','#5356FF','#FDA403','#FF204E')

w <- tmap_leaflet(
  tm_shape(pred[2,,]) +
    tm_raster(palette=pl2)
)

w

#saveWidget(w,here('R','output','bk_prob.html'))
```

### Estimate Nonnative Acres By Probability Bracket

```{r}
counts.df <- data.frame(matrix(ncol=6,nrow=0))
names(counts.df) <- c('boro','0-0.2','0.2-0.4','0.4-0.6','0.6-0.8','0.8-1')

pred.df <- as.data.frame(pred)
first <- sum(pred.df$prediction.nonnative>0 & pred.df$prediction.nonnative < 0.2,na.rm=TRUE)
second <- sum(pred.df$prediction.nonnative>=0.2 & pred.df$prediction.nonnative < 0.4,na.rm=TRUE)
third <- sum(pred.df$prediction.nonnative>=0.4 & pred.df$prediction.nonnative < 0.6,na.rm=TRUE)
fourth <- sum(pred.df$prediction.nonnative>=0.6 & pred.df$prediction.nonnative < 0.8,na.rm=TRUE)
fifth <- sum(pred.df$prediction.nonnative>=0.8 & pred.df$prediction.nonnative <= 1,na.rm=TRUE)

# function to get acres from 30m pixels
get_acres <- function(x){(x * 900)/4047}

new_row <- lapply(c(first,second,third,fourth,fifth),FUN=get_acres)
new_row <- c(boro,new_row)

counts.df <- rbind(counts.df,new_row)

# summarize mean and standard deviation of acres

counts.summary <- counts.df %>%
  group_by(boro) %>%
  summarize(mean.2 = mean(`0-0.2`),
         mean.4 = mean(`0.2-0.4`),
         mean.6 = mean(`0.4-0.6`),
         mean.8 = mean(`0.6-0.8`),
         mean.1 = mean(`0.8-1`),
         sd.2 = sd(`0-0.2`),
         sd.4 = sd(`0.2-0.4`),
         sd.6 = sd(`0.4-0.6`),
         sd.8 = sd(`0.6-0.8`),
         sd.1 = sd(`0.8-1`),
         mean.total = mean(total))
```


